---
title: "The _h5mread_ package"
author:
- name: Hervé Pagès
  affiliation: Fred Hutchinson Cancer Research Center, Seattle, WA
date: "Compiled `r doc_date()`; Modified 1 December 2024"
package: h5mread
vignette: |
  %\VignetteIndexEntry{The h5mread package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document
---


# Introduction

`r Biocpkg("h5mread")` is an R/Bioconductor package that allows fast and
memory-efficient loading of HDF5 data into R. The main function in the
package is `h5mread()` which allows reading arbitrary data from an HDF5
dataset into R, similarly to what the `h5read()` function from the
`r Biocpkg("rhdf5")`  package does. In the case of `h5mread()`, the
implementation has been optimized to make it as fast and memory-efficient
as possible.

In addition to the `h5mread()` function the package also provides the
following low-level functionality:

- `h5dim()` and `h5chunkdim()`;

- Utility functions to manipulate the dimnames of an HDF5 dataset;

- H5File objects to facilitate access to remote HDF5 files like files
  stored in an Amazon S3 bucket.

Note that the primary use case for the `r Biocpkg("h5mread")` package
is to support higher-level functionality implemented in
the `r Biocpkg("HDF5Array")` package.


# Install and load the package

Like any other Bioconductor package, `r Biocpkg("h5mread")` should always
be installed with `BiocManager::install()`:
```{r, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
BiocManager::install("h5mread")
```

Load the package:
```{r, message=FALSE}
library(h5mread)
```


# The h5mread() function

`h5mread()` is an efficient and flexible alternative to `rhdf5::h5read()`.

Note that we'll use `writeHDF5Array()` from the `r Biocpkg("HDF5Array")`
package to conveniently create the HDF5 datasets used in the examples below.

## Basic example

Create a 70,000 x 1,500 random dataset:
```{r, results='hide', message=FALSE}
set.seed(2009)
m0 <- matrix(runif(105e6), ncol=1500)  # 70,000 x 1,500 matrix
temp0_h5 <- tempfile(fileext=".h5")
HDF5Array::writeHDF5Array(m0, temp0_h5, "m0", chunkdim=c(100, 100))
```

Load 1,000 random rows from the HDF5 dataset:
```{r}
h5ls(temp0_h5)
nrow0 <- h5dim(temp0_h5, "m0")[[1L]]
starts <- list(sample(nrow0, 1000), NULL)
m <- h5mread(temp0_h5, "m0", starts=starts)
```

See `?h5mread` for more information and additional examples.

Sanity check:
```{r}
stopifnot(identical(m, m0[starts[[1L]], ]))
```

## An example involving sparse data

The HDF5 format doesn't natively support sparse data representation.
However, because it stores the data in compressed chunks, sparse data
gets efficiently compressed in the HDF5 file.

The more sparse the data, the smaller the resulting HDF5 file:
```{r, results='hide'}
a1 <- poissonSparseArray(c(6100, 960, 75), density=0.5)
a2 <- poissonSparseArray(c(6100, 960, 75), density=0.05)

temp1_h5 <- tempfile(fileext=".h5")
HDF5Array::writeHDF5Array(a1, temp1_h5, "a1", chunkdim=c(50, 40, 5), level=5)
temp2_h5 <- tempfile(fileext=".h5")
HDF5Array::writeHDF5Array(a2, temp2_h5, "a2", chunkdim=c(50, 40, 5), level=5)
```

```{r}
file.size(temp1_h5)
file.size(temp2_h5)
```

However, the small size on disk won't translate into a small size in memory
if the data gets loaded back as an ordinary array:
```{r}
a21 <- h5mread(temp2_h5, "a2")  # not memory-efficient
object.size(a21)
```

To keep memory usage as low as possible, `h5mread()` can load the data in
a SparseArray derivative from the `r Biocpkg("SparseArray")` package. This
is achieved by setting its `as.sparse` argument to `TRUE`:
```{r}
a22 <- h5mread(temp2_h5, "a2", as.sparse=TRUE)  # memory-efficient
object.size(a22)
```

Note that the data is loaded as a COO\_SparseArray object:
```{r}
class(a22)
```

See `?h5mread` for more information and additional examples.

Sanity checks:
```{r}
stopifnot(
  identical(a21, as.array(a2)),
  is(a22, "COO_SparseArray"),
  identical(as.array(a22), as.array(a2))
)
```

## A note about COO\_SparseArray objects

The COO\_SparseArray class is one of the two SparseArray concrete
subclasses defined in the `r Biocpkg("SparseArray")` package, the other
one being SVT\_SparseArray. Note that the latter tends to be even more
memory-efficient than the former and to achieve better performance overall.

Use coercion to switch back and forth between the two representations:
```{r}
a22 <- as(a22, "SVT_SparseArray")
class(a22)
```

About half the memory footprint of the COO\_SparseArray representation:
```{r}
object.size(a22)
```

See `?SparseArray` in the `r Biocpkg("SparseArray")` package for more
information.


# Other functionality provided by the h5mread package

## h5dim() and h5chunkdim()

Two convenience functions to obtain the dimensions of an HDF5
dataset as well as the dimensions of its chunks.

See `?h5dim` for more information and some examples.

## Utility functions to manipulate the dimnames of an HDF5 dataset

A small set of low-level utilities is provided to manipulate the
dimnames of an HDF5 dataset.

See `?h5writeDimnames` for more information and some examples.

## H5File objects

Use an H5File object to access an HDF5 file stored in an Amazon S3 bucket.

See `?H5File` for more information and some examples.


# Session information

```{r}
sessionInfo()
```

